Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-02-26 Latex/letter/letter.Rmd
txt updates

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefins in Bioinformatics, 2016). However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.

@@

> TO DO.

@@

> We thank the reviewer for the constructive comment. As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In fact, not all the analyzed marks in this paper are associated with the expression level of the neared gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. We believe that this analysis would be suboptimal to the one presented in this paper.

>>>>

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.

@@

> We thank the anonymous reviewer for this constructive comment. We note that ChromHMM benchmarked against the our method is the most widely used tool to perform chromatin segmentation analyses, and it is present as a standard method for large consortium projects such as ENCODE, Roadmap Epigenomics, Blueprint, CEEHRC, Mouse ENCODE, and Drosophila modENCODE^[https://dx.doi.org/10.1038%2Fnprot.2017.124]. In addition, we believe that the comparison of our method to the suggested algorithms would provide a secondary analysis for our paper due to the following reasons.

> First, jMOSAiCS is a package for chromatin segmentation that was removed from the Bioconductor project during release 3.4 in October 2016^[http://bioconductor.org/about/removed-packages/]. Second, methods have highlighted several limitations of dPCA^[https://dx.doi.org/10.1038%2Fncomms8973] in the analysis of chromatin segmentation, which includes the sensitivity to data scaling inherent to PCA, and the lack of covariate correction for batch and experiment-specific biases. Third, SignalSpider is an algorithm based on *Gaussian mixture models* for the clustering of enhancer and gene promoter regions across multiple ChIP-seq dat sets. We note that epigenomic marks associated with such regions are *narrow* in comparison to those analyzed in our paper, and their analysis via mixture models can be reasonably justified since read count enrichment is often present on narrow genomic positions. In contrast, our model is tailored for the analysis of marks for which read count enrichment expand through *broad* genomic domains. To this end, our model accounts for the local dependency of read enrichment from broad regions via *hidden Markov models* while still providing the combinatorial pattern detection via *negative binomial mixture models*. Hence, we believe that such a comparison would not be relevant for the readership of our article.

> In comparison to other HMM-based differential peak calling algorithms, we believe that the methodology presented in our paper provides several novelties over current approaches. First, the hidden Markov model presented in our article has an embedded mixture model in one of its components to model the combinatorial patterns of any number of conditions. Methods such as THOR and RSEG (both benchmarked in our paper), as well as the suggested histoneHMM, are only capable of analyzing two conditions at a time. Our ability to annotate combinatorial patterns of enrichment via mixture model posterior proportions is a novelty in comparison to HMM-based methods in both two- and multi-sample analyses. Second, our method provides an additional degree of flexibility in terms of data normalization for non linear biases as well as GC-content bias via model offsets in comparison to the aforementioned methods. Third, the implementation of our HMM is novel in the sense that we utilize an efficient rejection-controlled EM algorithm that leads to a fast estimation procedure and makes our algorithm among the fastest differential peak callers present in the literature. The suggested algorithm HPeak is a single-sample peak caller, in contrast to our multi-sample multi-condition differential peak caller, for which no software is available for download as of February 25 2020^[See https://sph.umich.edu/csg/qin/HPeak and https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-11-369].

@@

> We thank the reviewer for the constructive comment. As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene. In contrast, Figure 3 below presents an additional example of a DE gene and without H3K36me3 enrichment in both Helas3 and Hepg2 cell lines. Moreover, not all the analyzed marks in this paper are associated with the expression level of the nearest gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. In several instances, enrichment of the repressive mark H3K27me3 occurs on inter-genic regions with nearby expressed genes. Figure 4 below shows an example of such a region. Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.

![Third example of nonconcordant gene expression level and H3K36me3 ChIP-seq enrichment in Helas3. Here, a differentially expressed gene without differential erichment of H3K36me3](Fig_RNAseq_Screenshot3.pdf)

![Examples of H3K27me3 (a repressive mark) enrichment on intergenic regions. Note, on the right, an example of H3K27me3 enrichment on an intergenic region next to an expressed gene in both Helas3 and Hepg2 cell lines.](Fig_RNAseq_Screenshot4.pdf)
